Hilbert space methods to approximate Gaussian processes using Stan

### 2020-03-14 Paper/manuscript_v1/manuscript_06.tex
update in manuscript_v2 and online appendice

@@

\section{Introduction}\label{sec:bf_intro}

@@

In this paper we focus on the basis function approximation via Laplace eigenfunctions for stationary covariance functions proposed by \citet{solin2018hilbert}. Basis function approaches behave computationally like linear models, which is an attractive property in modular probabilistic programming models where there is a big benefit if approximation specific computation is simple (Paul: What do you mean with the the last part of the sentence?). The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance kernel including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost when only changing the covariance function parameters is $O(mn+m)$.% (or even $O(m^3)$ if marginalization can be used).
This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentation methods used in modern probabilistic programming frameworks (cite!). This is because the memory requirements of automatic differentation rather scale with the computational complexity instead of with the usual memory requirements for the posterior density computation. The basis function approach also provides an easy way to apply the non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which further improves MCMC efficiency.

@@

While \citet{solin2018hilbert} have fully developed the mathematical theory behind this specific approximation of GPs, further work is needed for its practical implementation in probabilistic programming frameworks such as Stan \citep{carpenter2017stan}. In this paper, we analyze in detail the performance and accuracy of the method in relation to key factors such as the number of basis functions, desired prediction space, or properties of the true functional relationship between covariates and response variable. We provide intuitive visualizations and practical recommendations for the choice of these factors, which will help users to improve computational performance while maintaining close approximation to exact GPs.

@@

\vspace{3mm}
\section{Method}\label{sec:bf_method}

@@

The mean and covariance functions completely characterize the Gaussian process prior, and control the a priori behavior of the function $f$. Let $\bm{f}=\left\lbrace f(\bm{x}_n) \right\rbrace_{n=1}^N$, then the resulting prior distribution for $\bm{f}$ is a multivariate Gaussian distribution
$\bm{f} \sim \mathcal{N}(\bm{\mu}, \bm{K})$,
 where $\bm{\mu} = \left\lbrace \mu(\bm{x}_n) \right\rbrace_{n=1}^N$ is the mean and $\bm{K}$ the covariance matrix, where $K_{i,j}=k(\bm{x}_i,\bm{x}_j)$, which needs be positive semidefinite. The covariance function $k(\bm{x}, \bm{x}')$ might depend on a set of hyperparameters, $\bm{\theta}$, but we will not write this dependency explicitly to ease the notation.
The joint distribution of $\bm{f}$ and a new $f^*$ is also a multivariate Gaussian as,
%
\begin{align*}
p(\bm{f}, f^*)=\mathcal{N} \left( \left[ \begin{array}{cc}
\bm{f} \\ 
f^*
\end{array} \right] \,\middle|\, \bm{0},\left[ \begin{array}{cc}
\bm{K}_{\bm{f},\bm{f}} & \bm{k}_{\bm{f},f^*} \\ 
\bm{k}_{f^*,\bm{f}} & k_{f^*,f^*}
\end{array} \right] \right),
\end{align*}

@@

The covariance function is the crucial ingredient in a Gaussian process as it encodes our prior assumptions about the function, and defines a correlation structure which characterize the correlations between function values at different inputs. A covariance function needs be symmetric and positive semidefinite. A stationary covariance function is a function of $\bm{\tau}=\bm{x}-\bm{x}' \in {\rm I\!R}^D$, such that it can be written $k(\bm{x},\bm{x}') = k(\bm{\tau})$, which means that the covariance is invariant to translations. Isotropic covariance functions are those that are function of the distance between observations, $k(\bm{x},\bm{x}') = k(|\bm{x}-\bm{x}'|) = k(r), r\in {\rm I\!R}$, which means that the covariance is both translation and rotation invariant. The most commonly used distance between observations is the norm L2 $(|\bm{x}-\bm{x}'|_{L2})$, also known as Euclidean distance, although other types of distances can be considered.

@@

\noindent in $D$ dimensions, where variable $\omega\in {\rm I\!R}$ is a distance in the frequency domain, and $\ell$ and $\sigma$ are the lengthscale and magnitud, respectively, of the kernel. The particular cases where $\nu=\infty$ and $\nu=3/2$ take the form
%
\begin{align}
S_{\infty}(\omega)&= \sigma^2 \sqrt{2\pi}^D  \ell^D  \exp\left(-0.5 \ell^2 \omega^2\right), \label{eq:specdens_inf}  \\
S_{\frac{3}{2}}(\omega)&= \sigma^2 \frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2})\sqrt{3}^3}{\frac{1}{2}\sqrt{\pi}\ell^3}\left(\frac{3}{\ell^2}+\omega^2 \right)^{-\frac{D+3}{2}}. \label{eq:specdens_32} 
\end{align}

@@

The approximate Gaussian process method, developed by \cite{solin2018hilbert} and implemented in this paper, is based on considering the covariance operator of a homogeneous (stationary) covariance function as a pseudo-differential operator constructed as a series of Laplace operators. Then, the pseudo-differential operator is approximated with Hilbert space methods on a compact subset $\Omega \subset {\rm I\!R}^D$ subject to some boundary condition. For brevity, we will refer to these approximate Gaussian processes as HSGPs. Below, we will present the main results around HSGPs relevant for practical application. More details and mathematical proofs are provided in \cite{solin2018hilbert}.

@@

Within $\Omega$, we can write any stationary covariance function with input values $\{x,x'\} \in \Omega$ as 
%
\begin{equation}\label{approxcov}
k(x,x') = \sum_{j=1}^\infty S_{\theta}(\sqrt{\lambda_j}) \phi_j(x) \phi_j(x'),
\end{equation}

\noindent where $S_{\theta}$ is the spectral density of the stationary covariance function $k$ (see Section 2.1) and $\theta$ the set of hyperparameters of $k$ \citep{rasmussen2006gaussian}. The terms $\{\lambda_j\}_{j=1}^{\infty}$ and $\{\phi_j(x)\}_{j=1}^{\infty}$ are the sets of eigenvalues and eigenvectors, respectively, of the Laplacian operator. Namely, they satisfy the following eigenvalue problem in $\Omega$ when applying the Dirichlet boundary condition (other boundary conditions could be used as well):
%
\begin{eqnarray}\label{eigenproblem}
\begin{split}
-\nabla^2 \phi_j(x)&=&\lambda \phi_j(x), \hspace{1cm}  x \in \Omega \\ 
\phi_j(x)&=&0, \hspace{2cm} x \notin \Omega.
\end{split}
\end{eqnarray}

\noindent The eigenvalues $\lambda_j>0$ are real and positive because the Laplacian is a positive definite Hermitian operator, and the eigenfunctions $\phi_j$ for the eigenvalues problem in Equation (\ref{eigenproblem}) are sinusoidal functions. Independently of the covariance function, they can be computed as
%
\begin{align}
\lambda_j&=\left(\frac{j\pi}{2L}\right)^2, \label{eigenvalue}\\
\phi_j(x)&=\sqrt{\frac{1}{L}} \text{sin}\left(\sqrt{\lambda_j}(x+L)\right). \label{eigenfunction}
\end{align}

If we truncate the sum in (\ref{approxcov}) to the first $m$ terms, the approximate covariance function becomes
%
\begin{equation}
k(x,x') \approx \sum_{j=1}^m S_{\theta}(\sqrt{\lambda_j}) \phi_j(x) \phi_j(x') = \bm{\phi}(x)^\intercal \Delta \bm{\phi}(x'), \nonumber
\end{equation}

@@

\noindent This equivalently leads to a linear representation of $f$ via
%
\begin{equation}\label{approxf}
f(x) \approx \sum_{j}^m \left( S_{\theta}(\sqrt{\lambda_j})\right)^{1/2} \phi_j(x) \beta_j,
\end{equation}

\noindent where $\beta_j \sim \text{Normal}(0,1)$. Thus, the function $f$ is approximated with a finite basis function expansion (using the eigenfunctions $\phi_j$ of the Laplace operator), scaled by the square root of spectral density values. A key property of this approximation is that the eigenfunctions $\phi_j$ do not depend on the covariance hyperparameters $\theta$. Instead, the only dependence on $\theta$ is through the spectral density $S_{\theta}$. The eigenvalues $\lambda_j$ are monotonically increasing with $j$ and $S_{\theta}$ goes rapidly to zero for bounded covariance functions. Therefore, Equation (\ref{approxf}) can be expected to be a good approximation for a finite number of $m$ terms in the series as long as the inputs values $x_i$ are not too close to the boundaries $-L$ and $L$ of $\Omega$. The computational cost of univariate HSGPs scales as $O(nm + m)$, where $n$ is the number of observations and $m$ the number of basis functions.

The parameterization in (\ref{approxf}) is naturally in the non-centered
parameterization form with independent prior distribution on
$\beta_j$, which makes the posterior inference easier.

@@

The results from the previous section can be generalizes to a multidimensional input space with compact regular domain $\Omega=[-L_1,L_1] \times \dots \times [-L_d,L_d]$ and Dirichlet boundary conditions. 
In a $D$-dimensional input space, the total number of eigenfunctions and eigenvalues in the approximation is equal to the number of $D$-tuples, that is possible combinations of univariate eigenfunctions over all dimensions. The number of $D$-tuples is given by 
%
\begin{align} \label{m_multi}
m^{\ast} = \prod_{d=1}^{D} m_d,
\end{align}

@@

Each multivariate eigenfunction $\phi^{\ast}_j$ corresponds to the product of the univariate eigenfunctions whose indices corresponds to the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$, and each multivariate eigenvalue $\bm{\lambda}^{\ast}_j$ is a $D$-vector with elements that are the univariate eigenvalues whose indices correspond to the elements of the $D$-tuple $\mathbb{S}_{j\bm{\cdotp}}$. Thus, for $\bm{x}=\{x_d\}_{d=1}^D \in \Omega$ and $j=1,\ldots,m^{\ast}$, we have 
%
\begin{align}
\phi^{\ast}_j(\bm{x}) &= \prod_{d=1}^{D} \phi_{\mathbb{S}_{jd}}(x_d) = \prod_{d=1}^{D} \sqrt{\frac{1}{L_d}} \text{sin}\left(\sqrt{\lambda_{\mathbb{S}_{jd}}}(x_d+L_d)\right) \label{eigenfunction_multi} \\
%
\bm{\lambda}^{\ast}_j &= \left\{ \lambda_{\mathbb{S}_{jd}} \right\}_{d=1}^D =  \left\{ \left(\tfrac{\pi \mathbb{S}_{jd}}{2L_d}\right)^2 \right\}_{d=1}^D. \label{eigenvalue_multi}
\end{align}

\noindent The approximate covariance function is then represented as
%
\begin{equation}\label{approxcov_multi}
k(\bm{x},\bm{x}') \approx \sum_{j=1}^{m^{\ast}} 
S^{\ast}_{\theta}\left(\sqrt{\bm{\lambda}^{\ast}_j}\right)
\phi^{\ast}_j(\bm{x}) \phi^{\ast}_j(\bm{x}'),
\end{equation}

\noindent where $S^{\ast}_{\theta}$ is the spectral density of the $D$-dimensional covariance function (see Section 2.1). We can now write the approximate series expansion of the multivariate function $f$ as,
%
\begin{equation}\label{approxf_multi}
f(\bm{x}) \approx \sum_{j=1}^{m^{\ast}} 
\left( S^{\ast}_{\theta} \left(\sqrt{\bm{\lambda}^{\ast}_j} \right)\right)^{1/2}
\phi^{\ast}_j(\bm{x}) \beta_j, 
\end{equation}

@@

\subsection{Learning hyperparameters and model inference}

\vspace{2mm}
$\bullet$ It has an attractive computational cost as this basically turns the regular GP model into a linear model.

\vspace{2mm}
$\bullet$ The design matrix of the proposed linear model, which is composed of a basis of Laplace eigenfunctions, can be computed analytically and does not depend on the hyperparameters of the model, then it has to be computed only once with $O(n+m)$ computational demands.

\vspace{2mm}
$\bullet$ The weights associated to the basis functions in this linear model is a $m$-dimensional vector ($m$ is the number of basis functions) and their computation is an operation with $O(m)$ computational demands. The weights depend on the hyperparameters, then they have to be computed in every step of the HMC sampling method.

\vspace{2mm}
$\bullet$ The linear model is computed with complexity $O(nm)$, computed in every step of the HMC sampling method.

\vspace{2mm}
$\bullet$ In a fully Bayesian inference framework using sampling methods, the proposed approximate GP model has a computational complexity of $O(nm+m)$ in every step of the HMC method. In addition, the computation of the automatic differentiation to compute the gradients in this linear model scales $O(n)$?, an operation that must be computed in every step of the HMC method.

\vspace{2mm}
$\bullet$ Using maximizing marginal likelihood methods, the proposed model has a overall complexity of $O(nm^2)$. After this, evaluating the marginal likelihood and marginal likelihood gradients is an $O(m^3)$ operation in every step of the optimizer. (Arno's paper, pag. 7)

\vspace{2mm}
$\bullet$ The parameter posterior distribution in this approximate GP model is $m$-dimensional ($m<<n$) which helps the use of GP priors as latent functions. especially when sampling methods for inference are used. GP prior as latent functions is needed in generalized models.

In regular GPs and other approximate GP models and Splines models these features do not have so nice properties:

\vspace{2mm}
$\bullet$ In a regular GPs, the main computational complexity comes from the inversion of the covariance matrix which is in general a $O(n^3)$ operation. This operation has to be computed at every step of the HMC or optimizer.

\vspace{2mm}
$\bullet$ In regular GPs, the parameter posterior distributions is $N$-dimensional. It is known that when $N$ is of medium or large size there is high correlation between the $N$-dimensional latent function and the hyperparameters of the GP prior.

\vspace{2mm}
$\bullet$ In conventional sparse GP approximations, although the rank of the GP is reduced considerably to the number of inducing points, this still needs to do the autodiff and covariance matrix inversion.

\vspace{2mm}
$\bullet$ The Splines models are also a sort of basis functions expansion model, then the computational demands are similar to that in this approach. However in Splines models the lengthscale hyperparameter tend to be fixed and then the fit is covered by the magnitude parameter. In that sense, Splines models tend to loose the useful interpretation of the lengthscale parameter.

\vspace{2mm}
$\bullet$ In addition, the computation of the automatic differentiation to compute gradients in this linear model scales $O(n)$, which is an operation that must be computed in every step of the HMC method.

\vspace{2mm}
$\bullet$ In a regular GP model the automatic differentiation to compute the gradients of the covariance function scales $O(n^2)$, the dimension of the covariance matrix, and the full inversion of the covariance matrix scales $O(n^3)$. This operation has to be computed at every step of the HMC.

\vspace{2mm}
$\bullet$ In a sparse GP approach based on inducing points, although the rank of the GP is reduced considerably to the number of inducing points, this still needs to do the autodiff and covariance matrix inversion.

\vspace{2mm}
$\bullet$ The Splines models are also a sort of basis functions expansion model, then the computational demands are similar to that in this approach.

@@

\subsection{Dependency on the number of basis functions and the boundary condition}

As explained in Section 2, the approximation of the covariance function is a series expansion of eigenfunctions and eigenvalues of the Laplace operator in a given domain $\Omega$, for instance in a one-dimensional input space $\Omega=[-L,L]\subset \mathbb{R}$:
%
\begin{equation}\label{diffcov}
k(\tau) = \sum_{j=1}^{\infty} S_{\theta} \left(\sqrt{\lambda_j} \right) \phi_j(\tau) \phi_j(0),  \nonumber
\end{equation}

\noindent where $L$ describes the boundary condition, $j$ is the index for the eigenfunctions and eigenvalues, and $\tau=x-x'$ is the difference between two covariate values $x$ and $x'$ in $\Omega$. The eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ are given in Equations (\ref{eigenvalue}) and (\ref{eigenfunction}) for the unidimensional case and in Equations (\ref{eigenvalue_multi}) and (\ref{eigenfunction_multi}) for the multidimensional case. The number of basis functions can be truncated at some finite positive value $m$ such that the difference between the densities of the exact and approximate covariance functions is less than a predefined threshold $\varepsilon > 0$:
%
\begin{eqnarray}\label{diff_covs}
\int k(\tau) \, d\tau - 
\int \sum_{j=1}^m S_{\theta}\left(\sqrt{\lambda_j} \right) \phi_j(\tau) \phi_j(0) \, d\tau < \varepsilon.
\end{eqnarray}

The finite number $m$ of basis functions in the approximation needed to satisfy Equation (\ref{diff_covs}) depends on the non-linearity of the function to be learned, that is on its lengthscale $\ell$, which constitutes a hyperparameter of the GP. The approximation also depends on the boundary $L$ (see Equations (\ref{eigenvalue}), (\ref{eigenfunction}), (\ref{eigenvalue_multi}) and (\ref{eigenfunction_multi})), which will affect its accuracy especially near the boundaries. As we will see later on, $L$ will also influence the number of basis functions required in the approximation. In the present paper, we will set $L$ an extension of the desired covariate input domain $\Psi$. Without loss of generality, we can assume $\Psi$ to be symmetric around zero, that is $\Psi=[-S,S] \subset \mathbb{R}$. We now define $L$ as
%
\begin{equation}\label{eq:boundary}
L=c \cdot S,
\end{equation}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 5mm 0mm 5mm 0mm, clip]{fig1_Post_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{fig1_Cov_J.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 27mm 32mm 122mm 10mm, clip]{fig1_legend.pdf}}
\caption{Mean posterior predictive functions (left) and covariance functions (right) of both the regular GP model and the HSGP model for different number of basis functions $m$, with the boundary factor fixed to a large enough value.}
  \label{fig1_Post_J}
\end{figure}

>>>>

%% Add a new subsection
\usepackage{titlesec}
\usepackage{hyperref}

\titleclass{\subsubsubsection}{straight}[\subsection]

\newcounter{subsubsubsection}[subsubsection]
\renewcommand\thesubsubsubsection{\thesubsubsection.\arabic{subsubsubsection}}
\renewcommand\theparagraph{\thesubsubsubsection.\arabic{paragraph}} % optional; useful if paragraphs are to be numbered

\titleformat{\subsubsubsection}
  {\normalfont\normalsize\bfseries}{\thesubsubsubsection}{1em}{}
\titlespacing*{\subsubsubsection}
{0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}

\makeatletter
\renewcommand\paragraph{\@startsection{paragraph}{5}{\z@}%
  {3.25ex \@plus1ex \@minus.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries}}
\renewcommand\subparagraph{\@startsection{subparagraph}{6}{\parindent}%
  {3.25ex \@plus1ex \@minus .2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries}}
\def\toclevel@subsubsubsection{4}
\def\toclevel@paragraph{5}
\def\toclevel@paragraph{6}
\def\l@subsubsubsection{\@dottedtocline{4}{7em}{4em}}
\def\l@paragraph{\@dottedtocline{5}{10em}{5em}}
\def\l@subparagraph{\@dottedtocline{6}{14em}{6em}}
\makeatother

\setcounter{secnumdepth}{4}
\setcounter{tocdepth}{4}
%%

@@

\section{Introduction}\label{sec_bf_intro}

@@

In this paper we focus on the basis function approximation via Laplace eigenfunctions for stationary covariance functions proposed by \citet{solin2018hilbert}. Basis function approaches behave computationally like linear models, which is an attractive property in modular probabilistic programming models where there is a big benefit if approximation specific computation is simple (Paul: What do you mean with the the last part of the sentence?). The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance kernel including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost when only changing the covariance function parameters is $O(mn+m)$.% (or even $O(m^3)$ if marginalization can be used).
This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentation methods used in modern probabilistic programming frameworks, such as Stan \citep{carpenter2017stan}, WinBUGS \citep{lunn2000winbugs} and others. This is because the memory requirements of automatic differentation rather scale with the computational complexity instead of with the usual memory requirements for the posterior density computation. The basis function approach also provides an easy way to apply the non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which further improves MCMC efficiency.

@@

While \citet{solin2018hilbert} have fully developed the mathematical theory behind this specific approximation of GPs, further work is needed for its practical implementation in probabilistic programming frameworks. In this paper, we analyze in detail the performance and accuracy of the method in relation to key factors such as the number of basis functions, desired prediction space, or properties of the true functional relationship between covariates and response variable. We provide intuitive visualizations and practical recommendations for the choice of these factors, which will help users to improve computational performance while maintaining close approximation to exact GPs.

@@

\vspace{3mm}
\section{Method}\label{sec_bf_method}

@@

The mean and covariance functions completely characterize the Gaussian process prior, and control the a priori behavior of the function $f$. Let $\bm{f}=\left\lbrace f(\bm{x}_n) \right\rbrace_{n=1}^N$, then the resulting prior distribution for $\bm{f}$ is a multivariate Gaussian distribution
$\bm{f} \sim \mathcal{N}(\bm{\mu}, \bm{K})$,
 where $\bm{\mu} = \left\lbrace \mu(\bm{x}_n) \right\rbrace_{n=1}^N$ is the mean and $\bm{K}$ the covariance matrix, where $K_{i,j}=k(\bm{x}_i,\bm{x}_j)$, which needs be symmetric and positive semi-definite. The covariance function $k(\bm{x}, \bm{x}')$ might depend on a set of hyperparameters, $\bm{\theta}$, but we will not write this dependency explicitly to ease the notation.
The joint distribution of $\bm{f}$ and a new $f^*$ is also a multivariate Gaussian as,
%
\begin{align*}
p(\bm{f}, f^*)=\mathcal{N} \left( \left[ \begin{array}{cc}
\bm{f} \\ 
f^*
\end{array} \right] \,\middle|\, \bm{0},\left[ \begin{array}{cc}
\bm{K}_{\bm{f},\bm{f}} & \bm{k}_{\bm{f},f^*} \\ 
\bm{k}_{f^*,\bm{f}} & k_{f^*,f^*}
\end{array} \right] \right),
\end{align*}

@@

The covariance function is the crucial ingredient in a Gaussian process as it encodes our prior assumptions about the function, and defines a correlation structure which characterize the correlations between function values at different inputs. A covariance function needs be symmetric and positive semi-definite. A stationary covariance function is a function of $\bm{\tau}=\bm{x}-\bm{x}' \in {\rm I\!R}^D$, such that it can be written $k(\bm{x},\bm{x}') = k(\bm{\tau})$, which means that the covariance is invariant to translations. Isotropic covariance functions are those that are function of the distance between observations, $k(\bm{x},\bm{x}') = k(|\bm{x}-\bm{x}'|) = k(r), r\in {\rm I\!R}$, which means that the covariance is both translation and rotation invariant. The most commonly used distance between observations is the norm L2 $(|\bm{x}-\bm{x}'|_{L2})$, also known as Euclidean distance, although other types of distances can be considered.

@@

\noindent in $D$ dimensions, where variable $\omega\in {\rm I\!R}$ is a distance in the frequency domain, and $\ell$ and $\sigma$ are the lengthscale and magnitud, respectively, of the kernel. The particular cases where $\nu=\infty$ and $\nu=3/2$ take the form
%
\begin{align}
S_{\infty}(\omega)&= \sigma^2 \sqrt{2\pi}^D  \ell^D  \exp\left(-0.5 \ell^2 \omega^2\right), \label{eq_specdens_inf}  \\
S_{\frac{3}{2}}(\omega)&= \sigma^2 \frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2})\sqrt{3}^3}{\frac{1}{2}\sqrt{\pi}\ell^3}\left(\frac{3}{\ell^2}+\omega^2 \right)^{-\frac{D+3}{2}}. \label{eq_specdens_32} 
\end{align}

@@

The approximate Gaussian process method, developed by \cite{solin2018hilbert} and implemented in this paper, is based on considering the covariance operator of a homogeneous (stationary) covariance function as a pseudo-differential operator constructed as a series of Laplace operators. Then, the pseudo-differential operator is approximated with Hilbert space methods on a compact subset $\Omega \subset {\rm I\!R}^D$ subject to some boundary condition. For brevity, we will refer to these approximate Gaussian processes as HSGPs. Below, we will present the main results around HSGPs relevant for practical application. More details and mathematical proofs are provided in \cite{solin2018hilbert}. Our starting point for presenting the method is the main result obtained by \cite{solin2018hilbert} of the definition of the covariance function as a series expansion of eigenvalues and eigenfunctions of the Laplacian operator. The mathematical details of this approximation have been briefly presented in the Appendix \ref{app_details_formulas_approx_covfun} of this work.

@@

Within $\Omega$, we can write any stationary covariance function with input values $\{x,x'\} \in \Omega$ as 
%
\begin{equation}\label{eq_approxcov}
k(x,x') = \sum_{j=1}^\infty S_{\theta}(\sqrt{\lambda_j}) \phi_j(x) \phi_j(x'),
\end{equation}

\noindent where $S_{\theta}$ is the spectral density of the stationary covariance function $k$ (see Section 2.1) and $\theta$ the set of hyperparameters of $k$ \citep{rasmussen2006gaussian}. The terms $\{\lambda_j\}_{j=1}^{\infty}$ and $\{\phi_j(x)\}_{j=1}^{\infty}$ are the sets of eigenvalues and eigenvectors, respectively, of the Laplacian operator in the given domain $\Omega$. Namely, they satisfy the following eigenvalue problem in $\Omega$ when applying the Dirichlet boundary condition (other boundary conditions could be used as well):
%
\begin{eqnarray}\label{eq_eigenproblem}
\begin{split}
-\nabla^2 \phi_j(x)&=&\lambda \phi_j(x), \hspace{1cm}  x \in \Omega \\ 
\phi_j(x)&=&0, \hspace{2cm} x \notin \Omega.
\end{split}
\end{eqnarray}

\noindent The eigenvalues $\lambda_j>0$ are real and positive because the Laplacian is a positive definite Hermitian operator, and the eigenfunctions $\phi_j$ for the eigenvalues problem in Equation (\ref{eq_eigenproblem}) are sinusoidal functions. Independently of the covariance function, they can be computed as
%
\begin{align}
\lambda_j&=\left(\frac{j\pi}{2L}\right)^2, \label{eq_eigenvalue}\\
\phi_j(x)&=\sqrt{\frac{1}{L}} \text{sin}\left(\sqrt{\lambda_j}(x+L)\right). \label{eq_eigenfunction}
\end{align}

If we truncate the sum in (\ref{eq_approxcov}) to the first $m$ terms, the approximate covariance function becomes
%
\begin{equation}
k(x,x') \approx \sum_{j=1}^m S_{\theta}(\sqrt{\lambda_j}) \phi_j(x) \phi_j(x') = \bm{\phi}(x)^\intercal \Delta \bm{\phi}(x'), \nonumber
\end{equation}

@@

\noindent This equivalently leads to a linear representation of $f$ via
%
\begin{equation}\label{eq_approxf}
f(x) \approx \sum_{j}^m \left( S_{\theta}(\sqrt{\lambda_j})\right)^{1/2} \phi_j(x) \beta_j,
\end{equation}

\noindent where $\beta_j \sim \text{Normal}(0,1)$. Thus, the function $f$ is approximated with a finite basis function expansion (using the eigenfunctions $\phi_j$ of the Laplace operator), scaled by the square root of spectral density values. A key property of this approximation is that the eigenfunctions $\phi_j$ do not depend on the covariance hyperparameters $\theta$. Instead, the only dependence on $\theta$ is through the spectral density $S_{\theta}$. The eigenvalues $\lambda_j$ are monotonically increasing with $j$ and $S_{\theta}$ goes rapidly to zero for bounded covariance functions. Therefore, Equation (\ref{eq_approxf}) can be expected to be a good approximation for a finite number of $m$ terms in the series as long as the inputs values $x_i$ are not too close to the boundaries $-L$ and $L$ of $\Omega$. The computational cost of univariate HSGPs scales as $O(nm + m)$, where $n$ is the number of observations and $m$ the number of basis functions.

The parameterization in (\ref{eq_approxf}) is naturally in the non-centered
parameterization form with independent prior distribution on
$\beta_j$, which makes the posterior inference easier.

@@

The results from the previous section can be generalizes to a multidimensional input space with compact regular domain $\Omega=[-L_1,L_1] \times \dots \times [-L_d,L_d]$ and Dirichlet boundary conditions. 
In a $D$-dimensional input space, the total number of eigenfunctions and eigenvalues in the approximation is equal to the number of $D$-tuples, that is possible combinations of univariate eigenfunctions over all dimensions. The number of $D$-tuples is given by 
%
\begin{align} \label{eq_m_multi}
m^{\ast} = \prod_{d=1}^{D} m_d,
\end{align}

@@

Each multivariate eigenfunction $\phi^{\ast}_j$ corresponds to the product of the univariate eigenfunctions whose indices corresponds to the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$, and each multivariate eigenvalue $\bm{\lambda}^{\ast}_j$ is a $D$-vector with elements that are the univariate eigenvalues whose indices correspond to the elements of the $D$-tuple $\mathbb{S}_{j\bm{\cdotp}}$. Thus, for $\bm{x}=\{x_d\}_{d=1}^D \in \Omega$ and $j=1,\ldots,m^{\ast}$, we have 
%
\begin{align}
\bm{\lambda}^{\ast}_j &= \left\{ \lambda_{\mathbb{S}_{jd}} \right\}_{d=1}^D =  \left\{ \left(\tfrac{\pi \mathbb{S}_{jd}}{2L_d}\right)^2 \right\}_{d=1}^D. \label{eq_eigenvalue_multi} \\
%
\phi^{\ast}_j(\bm{x}) &= \prod_{d=1}^{D} \phi_{\mathbb{S}_{jd}}(x_d) = \prod_{d=1}^{D} \sqrt{\frac{1}{L_d}} \text{sin}\left(\sqrt{\lambda_{\mathbb{S}_{jd}}}(x_d+L_d)\right) \label{eq_eigenfunction_multi}
\end{align}

\noindent The approximate covariance function is then represented as
%
\begin{equation}\label{eq_approxcov_multi}
k(\bm{x},\bm{x}') \approx \sum_{j=1}^{m^{\ast}} 
S^{\ast}_{\theta}\left(\sqrt{\bm{\lambda}^{\ast}_j}\right)
\phi^{\ast}_j(\bm{x}) \phi^{\ast}_j(\bm{x}'),
\end{equation}

\noindent where $S^{\ast}_{\theta}$ is the spectral density of the $D$-dimensional covariance function (see Section 2.1). We can now write the approximate series expansion of the multivariate function $f$ as,
%
\begin{equation}\label{eq_approxf_multi}
f(\bm{x}) \approx \sum_{j=1}^{m^{\ast}} 
\left( S^{\ast}_{\theta} \left(\sqrt{\bm{\lambda}^{\ast}_j} \right)\right)^{1/2}
\phi^{\ast}_j(\bm{x}) \beta_j, 
\end{equation}

@@

@@

\subsection{Dependency on the number of basis functions and the boundary condition} \label{sec_dependency}

As explained in Section 2, the approximation of the covariance function is a series expansion of eigenfunctions and eigenvalues of the Laplace operator in a given domain $\Omega$, for instance in a one-dimensional input space $\Omega=[-L,L]\subset \mathbb{R}$:
%
\begin{equation*}
k(\tau) = \sum_{j=1}^{\infty} S_{\theta} \left(\sqrt{\lambda_j} \right) \phi_j(\tau) \phi_j(0), 
\end{equation*}

\noindent where $L$ describes the boundary condition, $j$ is the index for the eigenfunctions and eigenvalues, and $\tau=x-x'$ is the difference between two covariate values $x$ and $x'$ in $\Omega$. The eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ are given in Equations (\ref{eq_eigenvalue}) and (\ref{eq_eigenfunction}) for the unidimensional case and in Equations (\ref{eq_eigenvalue_multi}) and (\ref{eq_eigenfunction_multi}) for the multidimensional case. The number of basis functions can be truncated at some finite positive value $m$ such that the difference between the densities of the exact and approximate covariance functions is less than a predefined threshold $\varepsilon > 0$:
%
\begin{eqnarray}\label{eq_diff_covs}
\int k(\tau) \, d\tau - 
\int \sum_{j=1}^m S_{\theta}\left(\sqrt{\lambda_j} \right) \phi_j(\tau) \phi_j(0) \, d\tau < \varepsilon.
\end{eqnarray}

The finite number $m$ of basis functions in the approximation needed to satisfy Equation (\ref{eq_diff_covs}) depends on the non-linearity of the function to be learned, that is on its lengthscale $\ell$, which constitutes a hyperparameter of the GP. The approximation also depends on the boundary $L$ (see Equations (\ref{eq_eigenvalue}), (\ref{eq_eigenfunction}), (\ref{eq_eigenvalue_multi}) and (\ref{eq_eigenfunction_multi})), which will affect its accuracy especially near the boundaries. As we will see later on, $L$ will also influence the number of basis functions required in the approximation. In the present paper, we will set $L$ an extension of the desired covariate input domain $\Psi$. Without loss of generality, we can assume $\Psi$ to be symmetric around zero, that is $\Psi=[-S,S] \subset \mathbb{R}$. We now define $L$ as
%
\begin{equation}\label{eq_boundary}
L=c \cdot S,
\end{equation}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{fig1_Post_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{fig1_Cov_J.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 27mm 32mm 122mm 10mm, clip]{fig1_legend.pdf}}
\caption{Mean posterior predictive functions (left) and covariance functions (right) of both the regular GP model and the HSGP model for different number of basis functions $m$, with the boundary factor fixed to a large enough value.}
  \label{fig1_Post_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{fig2_Post_L.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{fig2_Cov_L.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 27mm 32mm 120mm 0mm, clip]{fig2_legend.pdf}}
\caption{Mean posterior predictive functions (left) and covariance functions (right) of both the regular GP model and the HSGP model for different values of the boundary factor $c$, with a large enough fixed number of basis functions.}
  \label{fig2_Post_L}
\end{figure}
